---
title: "Chap11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Case study

We study the distribution of the vegetation of Carpathian wetlands . The occurrences about vegetation of $30$ bryophytes plants together with environmental variables, recorded in $16 \ \text{m}^2$ plots from 70 fens (1 plot per locality) in the borderland between the Czech Republic and Slovakia. Water samples for consequent chemical analysis were taken in microsites best supplied by water from the central part of the fen, to obtain 14 chemical parameters of water ($Ca$, $Mg$, $Fe$, $K$, $Na$, $Si$, $SO_4$, $PO_4$, $NO_3$, $NH_3$, $Cl$, $Corg$, $pH$ and conductivity).


```{r download,results='hide', warning=FALSE,message = FALSE}
rm(list=ls())
library(factoextra)
library(Hmsc)
library(knitr)
library(corrplot)

#data download
bryo <- read.delim ('https://raw.githubusercontent.com/zdealveindy/anadat-r/master/data/bryophytes.txt', row.names = 1)
vasc <- read.delim ('https://raw.githubusercontent.com/zdealveindy/anadat-r/master/data/vasc_plants.txt', row.names = 1)
chem <- read.delim ('https://raw.githubusercontent.com/zdealveindy/anadat-r/master/data/chemistry.txt', row.names = 1)

X<-as.data.frame(apply(chem[,1:14],MARGIN=2, scale))
attach(X)

set.seed(17121993)

# info of data here: https://www.davidzeleny.net/anadat-r/doku.php/en:data:wetlands
```

## Select variables

We apply a PCA to select the predictors. We choose $Corg$, $Mg$, $K$, $conduct$ after the PCA analyisis. We split the dataset to obtain a train and a test dataset ($70 \%$ and $30\%$ respectively).

```{r var_selection, echo=FALSE}
x.pca <- princomp(X)

#We select 5 varibles
fviz_eig(x.pca)


#we might also take Corg, Mg, K, conduct
fviz_pca_var(x.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
)

train <- sample(1:nrow(X), floor(0.7*nrow(X)),replace=F)

X_test <- X[-train,]
X_train <- X[train,]

Y <- as.matrix(bryo)
Y_test <- Y[-train,]
Y_train <- Y[train,]

```

## Model definition

We include the quadratic terms of each variable to allow for local optima. We choose to model the response data with the Poisson distribution since we are modelling counts.

```{r ,model_def, echo=TRUE}
XFormula = ~ poly(Corg,degree=2) + poly(Mg,degree=2) + poly(K,degree=2) + poly(conduct, degree=2)

studyDesign = data.frame(sample = as.factor(1:nrow(Y_train))) 
rL = HmscRandomLevel(units = studyDesign$sample)

m = Hmsc(Y = Y_train,
       XData = X_train, XFormula = XFormula, ,ranLevels = list(sample = rL), studyDesign = studyDesign,
       distr = "poisson")
```

## MCMC run
We run two chains of the MCMC sampler, thinned to keep one every 10 samples, with 1000 elements of burn-in for a sample size of 20 00 .
```{r ,mcmc, echo=TRUE, results='hide'}
thin = 10
samples = 1000
nChains = 2

m = sampleMcmc(m, samples = samples, thin = thin,
               transient = 1000, nChains = nChains,verbose=1000, nParallel = nChains)
```


## Convergence analysis

We evaluate MCMC convergence in terms of the parameters that we are especially interested in:  the species niches, given by the elements in matrix $\boldsymbol{B}$, and the residual species associations $\boldsymbol{Sigma}$. Globally the diagnosis shows a good MCMC convergence, as most potential scale reduction factors are close to one and effective sample sizes are relatively high.

```{r ,conv_analysis, echo=TRUE}
mpost = convertToCodaObject(m)
par(mfrow=c(2,2))
hist(effectiveSize(mpost$Beta), main="ess(beta)",xlab="") 
hist(gelman.diag(mpost$Beta,multivariate=FALSE)$psrf, main="psrf(beta)",xlab="") 
hist(effectiveSize(mpost$Omega[[1]]), main="ess(omega)",xlab="") 
hist(gelman.diag(mpost$Omega[[1]],multivariate=FALSE)$psrf, main="psrf(omega)",xlab="")

```

## Posterior estimates: the regression coefficients

We represent here the response of each species to the environement. We only show the coefficients whose credible interval does not overlap zero.
```{r , beta_est , echo=TRUE}

postBeta = getPostEstimate(m, parName="Beta")
plotBeta(m, post=postBeta, param="Support", supportLevel = 0.95,cex = c(.35,.4,.5))
```

## Posterior estimates: species associations matrix

The species associations matrix is represented here, by showing only those values of the association matrix whose credible intervals do not overlap zero.
```{r , omega_est , echo=TRUE}
OmegaCor = computeAssociations(m)
supportLevel = 0.95

toPlot = ((OmegaCor[[1]]$support>supportLevel)
          + (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot(toPlot, method = "color", col=colorRampPalette(c("blue","white","red"))(200), title=paste("Species associations matrix"), mar=c(0,0,1,0), type="lower",order = "hclust", tl.cex=0.5)

```

## Ordination

We represent the species in a biplot whose axis are the two first latent factors. Species that are closer together are more correlated, and share the same response to the missing environmental variables that are represented by the values of the  

```{r , ordination , echo=TRUE}
etaPost = getPostEstimate(m, "Eta")
lambdaPost=getPostEstimate(m, "Lambda")
biPlot(m, etaPost = etaPost, lambdaPost=lambdaPost, factors=c(1,2))
```

## Model fit

We evaluate model fit in terms of $R^2$ both on the training set and on cross validation on the training set.

```{r , prediction , echo=TRUE, message=FALSE, warning=FALSE,results='hide'}
preds = computePredictedValues(m)
MF = evaluateModelFit(hM=m, predY=preds)

partition = createPartition(m, nfolds = 2)
preds = computePredictedValues(m, partition = partition, verbose=FALSE)
CV = evaluateModelFit(hM = m, predY = preds)

studyDesign_new = data.frame(sample = as.factor(1:nrow(X_test)))
rL_new = HmscRandomLevel(units = studyDesign_new$sample) 
#predict unconditional on test 
Uncond=predict(m, XData= X_test, study=studyDesign_new, ranLevels = list(sample=rL_new), expected=F) 
meanUncond = apply(simplify2array(Uncond), 1:2,mean) 

pred_test = sqrt(colSums((meanUncond-Y_test)^2)/nrow(Y_test))


```
```{r , prediction _figure, echo=FALSE, message=FALSE, warning=FALSE}
ggp = data.frame(value = c(MF$RMSE,CV$RMSE, pred_test),type=c(rep("Train",length(MF$C.RMSE)),rep("Cross Validation",length(CV$C.RMSE)), rep("Test", length(pred_test))))

ggplot()+aes(x=ggp$value, fill=as.factor(ggp$type))+
    geom_histogram(color="#e9ecef", alpha=0.4, position = 'identity') +
    scale_fill_manual(values=c("#69b3a2", "#404080","#FF6666")) +
    #theme_ipsum() +
  
    labs(fill="", xlabs=" ",title="RMSE")

table = data.frame("Train RMSE"= mean(MF$RMSE), "CV RMSE" = mean(CV$RMSE), "pred RMSE"= mean(pred_test))
kable(table)
#to ggplot with two histograms superposed
```

